Modelling BCG vaccination in the UK

### 2019-08-06 rmds/07-direct-effect-bcg-change.Rmd
update word count
chapter 7 formatting
edited up to chapter 8

## Read formated results
## See https://github.com/seabbs/DirectEffBCGPolicyChange/blob/master/vignettes/paper.Rmd
## for code required to reproduce 
## Chapter is an edited version of the orginal reproducible document.
##Wordcount: 4410
read_formated <- function(name) {
  path <- file.path("chapters/direct-eff-bcg-change", paste0(name, ".rds"))

@@

Globally, several countries with low TB incidence have moved from universal vaccination, either of those at school-age or neonates, to targeted vaccination of neonates considered at high-risk of TB  (see Chapter \@ref(background)).[@Zwerling2011a] In Sweden, which discontinued universal vaccination of neonates in favour of targeted vaccination of those at high risk, incidence rates in Swedish-born children increased slightly after the change in policy. [@Romanus1992] In France, which also switched from universal vaccination of neonates to targeted vaccination of those at high-risk, a study found that targeted vaccination of neonates may have reduced coverage in those most at risk.[@Guthmann2011]

@@

R `3.5.0` was used for all analysis.[@R] Reproducibility was ensured by using `R` package infrastructure^[Analysis R package DOI: [10.5281/zenodo.2551554](https://zenodo.org/badge/latestdoi/93072437)]. Missing data imputation using a GBM was implemented using the `h2o` package (see Section \@ref(imp-uk-birth-status)).[@h2o2018] Incidence rates, with 95% confidence intervals, were calculated using the `epiR` package (see Chapter \@ref(data)).[@EpiR] The `brms` package,[@Burkner] and STAN,[@StanDevelopmentTeam2016] was used to perform MCMC. Models were run until convergence (4 chains with a burn in of 10,000, and 10,000 sampled iterations each), with convergence being assessed using trace plots and the R hat diagnostic.[@StanDevelopmentTeam2016] All numeric confounders were centered and scaled by their standard deviation, and age was adjusted for using single year of age categories.

@@

As I was imputing a single variable I reformulated the imputation as a categorical prediction problem. This allowed the use of more complex, high-performing models compared to those usually used for imputation, whilst also allowing the results to be validated it using predictive modelling performance metrics. I included year of notification, sex, age, PHE Centre (PHEC), occupation, ethnic group, Index of Multiple Deprivation (2010) categorised into five groups for England (IMD rank), and risk factor count (risk factors considered; drug use, homelessness, alcohol misuse/abuse and prison). However, I could not account for a possible missing not at random mechanism not captured by these covariates. To train the model I first split the data with complete UK birth status into a training set (80%), a calibration set (5%), and a test set (15%). I then fit a gradient boosted machine with 10,000 trees, early stopping (at a precision of $1e-5$, with 10 stopping rounds), a learning rate of 0.1, and a learn rate annealing of 0.99. Gradient boosted machines are a tree based method that can incorporate complex non-linear relationships and interactions. Much like a random forest model they work by ensembling a group of trees, but unlike a random forest model each tree is additive aiming to reduce the residual loss from previous trees. Once the model had been fit to the training set I performed platt scaling using the calibration data set. Our fitted imputation model had a Logloss of 0.28 on the test set, with an AUC of 0.93, both of which indicate a robust out of bag performance. I found that ethnic group was the most important variable for predicting UK birth status, followed by age and PHEC.

@@

```{r 07-imp-com}
read_formated("imputation-evaluation") %>% 
  kable(caption = "Comparison of UK birth status in cases with complete or imputed records.", booktabs = TRUE) %>% 
  kable_styling(latex_options = "hold_position")
```

Inclusion of imputed values for UK birth status should reduce bias caused by any missing not at random mechanism captured by predictors included in the model. Graphical evaluation of UK birth status indicated that missingness has reduced over time, indicating a missing not at random mechanism (see Chapter \@ref(data)). If only the complete case data had been included in the analysis then incidence rates would have reduced over the study period due to this mechanism, this may have biased the estimate of the impact of the change in policy.

@@

```{r 07-cohort-summary-plot, fig.scap = "Mean incidence rates per 100,000, with 95\\% confidence intervals for each retrospective cohort, stratified by the vaccination policy and UK birth status.", fig.cap = "Mean incidence rates per 100,000, with 95\\% confidence intervals for each retrospective cohort (see Table \\\\@ref(tab:07-cohorts) for cohort definitions), stratified by the vaccination policy and UK birth status. The top and bottom panels are on different scales in order to highlight trends in incidence rates over time.", out.extra = ""}
knitr::include_graphics("chapters/direct-eff-bcg-change/figure-1.png")
```

Trends in incidence rates varied by age group and UK birth status. From 2000 until 2012 incidence rates in the UK born remained relatively stable but have since fallen year on year. In comparison incidence rates in the non-UK born increased from 2000 until 2005, since when they have also decreased year on year (see Chapter \@ref(data)). In 14-19 year old’s, who were UK born, incidence rates remained relatively stable throughout the study period, except for the period between 2006 to 2009 in which they increased year on year. This trend was not observed in the non-UK born population aged 14-19, where incidence rates reached a peak in 2003, since when they have consistently declined. In those aged 0-5, who were UK born, incidence rates also increased year on year after the change in BCG policy, until 2008 since when they have declined. This does not match with the observed trend in incidence rates in the non-UK born population, aged 0-5, in which incidence rates declined steeply between 2005 and 2006, since when they have remained relatively stable (Figure \@ref(fig:07-inc-rate-summary-plot)).

@@

For the comparable non-UK born cohort who were relevant to the targeted neonatal vaccination programme there was evidence, across all models, that change in policy was associated with a large decrease in incidence rates (IRR: `r extract_inc_ef(ModelComTargetedNonUKBorn, ModelComTargetedNonUKBorn$Model[1])`) (Table \@ref(tab:07-targeted-tab) in the best fitting model). The best fitting model was a Negative Binomial model that adjusted for the change in policy, age, and non-UK born incidence rates (Table \@ref(tab:07-targeted-tab)). All models which at least adjusted for age estimated comparable effects of the change in policy (Table \@ref(tab:07-summary-targeted-nonukborn)).

@@

I was unable to estimate the impact of the change in BCG policy after 5 years post vaccination, so both the estimates of the positive and negative consequences are likely to be underestimates of the ongoing impact. TB is a complex disease and the BCG vaccine is known to offer imperfect protection, which has been shown to vary both spatially and with time since vaccination (see Chapter \@ref(background)).[@Mangtani2014a; @Abubakar2013] By focusing on the impact of the change in policy on the directly affected populations within a short period of time, and by employing a multi-model approach I have limited the potential impact of these issues. This study was based on a routine observational data set (ETS), and a repeated survey (LFS) both of which may have introduced bias. Whilst the LFS is a robust data source, widely used in academic studies,[@French2007; @Davies2016a; @Lindley2009] it is susceptible to sampling errors particularly in the young, and in the old, which may have biased the estimated incidence rates. As the ETS is routine surveillance system some level of missing data is inevitable (see Chapter \@ref(data)). However, UK birth status is relatively complete (93% (106765/114820)) and I imputed missing values using an approach which accounted for MNAR mechanisms for the variables included in the imputation model. I was unable to adjust for known demographic risk factors for TB, notably socio-economic status,[@Parslow2001; @Bhatti1995] and ethnicity.[@Parslow2001; @Bhatti1995; @Abubakar2008] However, this confounding is likely to be mitigated by the use of multiple cohorts and the adjustment for incidence rates in the UK born and non-UK born. Finally, I have assumed that the effect I have estimated for the change in BCG policy is due to the changes in BCG vaccination policy as well as other associated changes in TB control policy, after adjusting for hypothesised confounders. However, there may have been additional policy changes which I have not accounted for.

@@

It is well established that interventions against infectious diseases, such as TB, should be evaluated not only for their direct effects but also for future indirect effects via ongoing transmission. Statistical approaches such as those used in this paper are not appropriate for capturing these future indirect effects, and instead dynamic disease models should be used. In Chapter \@ref(model-development) I develop such a dynamic disease model, Chapter \@ref(model-fitting) then fits this model to the available data, and Chapter \@ref(model-results) compares the impact of continuing with the BCG school's scheme post 2005 compared to universal neonatal vaccination. In addition, this study could not evaluate the impact of the neonatal programme on the high-risk population it targets, due to a lack of reliable data. Improved coverage data for the BCG programme is required to more fully evaluate its ongoing impact.

@@

* Overall, we found that the change in BCG policy was associated with directly preventing 385 (95% CI -105, 881) TB cases.

@@

* The code for the analysis contained in this Chapter can be found at: [doi.org/10.5281/zenodo.2583056](doi.org/10.5281/zenodo.2583056)^[Alternatively available from: https://github.com/seabbs/DirectEffBCGPolicyChange]

>>>>

## Read formated results
## See https://github.com/seabbs/DirectEffBCGPolicyChange/blob/master/vignettes/paper.Rmd
## for code required to reproduce 
## Chapter is an edited version of the orginal reproducible document.
##Wordcount: 4504
read_formated <- function(name) {
  path <- file.path("chapters/direct-eff-bcg-change", paste0(name, ".rds"))

@@

Globally, several countries with low TB incidence have moved from universal vaccination, either of those at school-age or neonates, to targeted vaccination of neonates considered at high-risk of TB  (see Chapter \@ref(background)).[@Zwerling2011a] In Sweden, which discontinued universal vaccination of neonates in favour of targeted vaccination of those at high risk, incidence rates in Swedish-born children increased slightly after the change in policy.[@Romanus1992] In France, which also switched from universal vaccination of neonates to targeted vaccination of those at high-risk, a study found that targeted vaccination of neonates may have reduced coverage in those most at risk.[@Guthmann2011]

@@

R `3.5.0` was used for all analysis.[@R] Reproducibility was ensured by using `R` package infrastructure^[Code: https://github.com/seabbs/DirectEffBCGPolicyChange]. Missing data imputation using a GBM was implemented using the `h2o` package (see Section \@ref(imp-uk-birth-status)).[@h2o2018] Incidence rates, with 95% confidence intervals, were calculated using the `epiR` package (see Chapter \@ref(data)).[@EpiR] The `brms` package,[@Burkner] and STAN,[@StanDevelopmentTeam2016] was used to perform MCMC. Models were run until convergence (4 chains with a burn in of 10,000, and 10,000 sampled iterations each), with convergence being assessed using trace plots and the R hat diagnostic.[@StanDevelopmentTeam2016] All numeric confounders were centered and scaled by their standard deviation, and age was adjusted for using single year of age categories.

@@

As I was imputing a single variable I reformulated the imputation as a categorical prediction problem. This allowed the use of more complex, high-performing models compared to those usually used for imputation, whilst also allowing the results to be validated using predictive modelling performance metrics. I included year of notification, sex, age, PHE Centre (PHEC), occupation, ethnic group, Index of Multiple Deprivation (2010) categorised into five groups for England (IMD rank), and risk factor count (risk factors considered; drug use, homelessness, alcohol misuse/abuse and prison). However, I could not account for a possible missing not at random mechanism not captured by these covariates. To train the model I first split the data with complete UK birth status into a training set (80%), a calibration set (5%), and a test set (15%). I then fit a gradient boosted machine with 10,000 trees, early stopping (at a precision of $1e-5$, with 10 stopping rounds), a learning rate of 0.1, and a learn rate annealing of 0.99. Gradient boosted machines are a tree based method that can incorporate complex non-linear relationships and interactions. Much like a random forest model they work by ensembling a group of trees, but unlike a random forest model each tree is additive aiming to reduce the residual loss from previous trees. Once the model had been fit to the training set I performed platt scaling using the calibration dataset. Our fitted imputation model had a Logloss of 0.28 on the test set, with an AUC of 0.93, both of which indicate a robust out of bag performance. I found that ethnic group was the most important variable for predicting UK birth status, followed by age and PHEC.

@@

```{r 07-imp-com}
read_formated("imputation-evaluation") %>% 
  kable(caption = "Comparison of UK birth status in cases with complete or imputed records.", booktabs = TRUE) %>% 
  kable_styling(latex_options = "HOLD_position")
```

Inclusion of imputed values for UK birth status should reduce bias caused by any missing at random mechanism captured by predictors included in the model. Graphical evaluation of UK birth status indicated that missingness has reduced over time, indicating a missing at random mechanism (see Chapter \@ref(data)). If only the complete case data had been included in the analysis then incidence rates would have reduced over the study period due to this mechanism, this may have biased the estimate of the impact of the change in policy.

@@

```{r 07-cohort-summary-plot, fig.scap = "Mean incidence rates per 100,000, with 95\\% confidence intervals for each retrospective cohort, stratified by the vaccination policy and UK birth status.", fig.cap = "Mean incidence rates per 100,000, with 95\\% confidence intervals for each retrospective cohort, stratified by the vaccination policy and UK birth status. The top and bottom panels are on different scales in order to highlight trends in incidence rates over time.", out.extra = ""}
knitr::include_graphics("chapters/direct-eff-bcg-change/figure-1.png")
```

Trends in incidence rates varied by age group and UK birth status. From 2000 until 2012 incidence rates in the UK born remained relatively stable but have since fallen year on year. In comparison, incidence rates in the non-UK born increased from 2000 until 2005, since when they have also decreased year on year (see Chapter \@ref(data)). In 14-19 year old’s, who were UK born, incidence rates remained relatively stable throughout the study period, except for the period between 2006 to 2009 in which they increased year on year. This trend was not observed in the non-UK born population aged 14-19, where incidence rates reached a peak in 2003, since when they have consistently declined. In those aged 0-5, who were UK born, incidence rates also increased year on year after the change in BCG policy, until 2008 since when they have declined. This does not match with the observed trend in incidence rates in the non-UK born population, aged 0-5, in which incidence rates declined steeply between 2005 and 2006, since when they have remained relatively stable (Figure \@ref(fig:07-inc-rate-summary-plot)).

@@

For the comparable non-UK born cohort who were relevant to the targeted neonatal vaccination programme there was evidence, across all models, that the change in policy was associated with a large decrease in incidence rates (IRR: `r extract_inc_ef(ModelComTargetedNonUKBorn, ModelComTargetedNonUKBorn$Model[1])`) (Table \@ref(tab:07-targeted-tab) in the best fitting model). The best fitting model was a Negative Binomial model that adjusted for the change in policy, age, and non-UK born incidence rates (Table \@ref(tab:07-targeted-tab)). All models which at least adjusted for age estimated comparable effects of the change in policy (Table \@ref(tab:07-summary-targeted-nonukborn)).

@@

I was unable to estimate the impact of the change in BCG policy after 5 years post vaccination, so both the estimates of the positive and negative consequences are likely to be underestimates of the ongoing impact. TB is a complex disease and the BCG vaccine is known to offer imperfect protection, which has been shown to vary both spatially and with time since vaccination (see Chapter \@ref(background)).[@Mangtani2014a; @Abubakar2013] By focusing on the impact of the change in policy on the directly affected populations within a short period of time, and by employing a multi-model approach I have limited the potential impact of these issues. This study was based on a routine observational dataset (ETS), and a repeated survey (LFS) both of which may have introduced bias. Whilst the LFS is a robust data source, widely used in academic studies,[@French2007; @Davies2016a; @Lindley2009] it is susceptible to sampling errors particularly in the young, and in the old, which may have biased the estimated incidence rates. As the ETS is routine surveillance system some level of missing data is inevitable (see Chapter \@ref(data)). However, UK birth status is relatively complete (93% (106765/114820)) and I imputed missing values using an approach which accounted for MAR mechanisms for the variables included in the imputation model. I was unable to adjust for known demographic risk factors for TB, notably socio-economic status,[@Parslow2001; @Bhatti1995] and ethnicity.[@Parslow2001; @Bhatti1995; @Abubakar2008] However, this confounding is likely to be mitigated by the use of multiple cohorts and the adjustment for incidence rates in the UK born and non-UK born. Finally, I have assumed that the effect I have estimated for the change in BCG policy is due to the changes in BCG vaccination policy as well as other associated changes in TB control policy, after adjusting for hypothesised confounders. However, there may have been additional policy changes which I have not accounted for.

@@

It is well established that interventions against infectious diseases, such as TB, should be evaluated not only for their direct effects but also for future indirect effects via ongoing transmission. Statistical approaches such as those used in this paper are not appropriate for capturing these future indirect effects, and instead dynamic disease models should be used. In Chapter \@ref(model-development) I develop such a dynamic disease model, Chapter \@ref(model-fitting) then fits this model to the available data, and Chapter \@ref(model-results) compares the impact of continuing with the BCG school's scheme post 2005 to universal neonatal vaccination. In addition, this study could not evaluate the impact of the neonatal programme on the high-risk population it targets, due to a lack of reliable data. Improved coverage data for the BCG programme is required to more fully evaluate its ongoing impact.

@@

* Overall, I found that the change in BCG policy was associated with directly preventing 385 (95% CI -105, 881) TB cases.

@@

* The code for the analysis contained in this Chapter can be found at: [doi.org/10.5281/zenodo.2583056](doi.org/10.5281/zenodo.2583056)^[Alternatively code link: https://github.com/seabbs/DirectEffBCGPolicyChange]
